---
title: "Project 2: Linear Regression"
author:
  - Jongbum Won (01815150)
output: html_document
date: '2022-04-21'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages("palmerpenguins")
```

```{r}
require(palmerpenguins)
penguins <- palmerpenguins::penguins
penguins = penguins[complete.cases(penguins),]
attach(penguins)
require(ggplot2)

```

# **1. Statistical Analysis**
### (a) Explanatory analysis
```{r}
summary(penguins)
```
Load penguins dataset.

```{r}
ggplot(penguins, 
       mapping=aes(bill_length_mm, body_mass_g, color=species, shape=species)) + 
  geom_smooth(method = "lm", level = 0.95 , formula = y ~ x, color='black', size = 0.7) +
  geom_smooth(method = "loess", level = 0.95 , se = FALSE, linetype=6, formula = y ~ x, size=0.5, color='black') +
  ggtitle('Body Mass (g) vs Bill Length (mm)') + geom_point()

ggplot(penguins, 
       mapping=aes(bill_depth_mm, body_mass_g, color=species, shape=species)) + 
  geom_smooth(method = "lm", level = 0.95 , formula = y ~ x, color='black', size = 0.7) +
  geom_smooth(method = "loess", level = 0.95 , se = FALSE, linetype=6, formula = y ~ x, size=0.5, color='black') +
  ggtitle('Body Mass (g) vs Bill Depth (mm)') + geom_point()

ggplot(penguins, 
       mapping=aes(flipper_length_mm, body_mass_g, color=species, shape=species)) + 
  geom_smooth(method = "lm", level = 0.95 , formula = y ~ x, color='black', size = 0.7) +
  geom_smooth(method = "loess", level = 0.95 , se = FALSE, linetype=6, formula = y ~ x, size=0.5, color='black') +
  ggtitle('Body Mass (g) vs Flipper Length (mm)') + geom_point()

```

Check if there is linear relatioship three continuous features (`bill_length_mm`, `bill_depth_mm`, and `filpper_length_mm`) and weight of an individual penguin (`body_mass_g`).  
Draw best fit linear regression lines (solid) together with loess (locally weighted scatterplot smoothing) curves (dotted) on the scatter plot for each species.  
The curves quite approximate well corresponding lines, except for ends of lines with low number of observations.  
Thus, from the graphs, we can conclude that it is reasonable to assume that there are linear relationships between `bill_length_mm` and `body_mass_g`, `bill_depth_mm` and `body_mass_g`, and `filpper_length_mm` and`body_mass_g`.  

Q Lowess vs loess? 

### (b) Building optimal regression model
1st Round
```{r}
model1 <- lm(body_mass_g ~ bill_length_mm) 
summary(model1)
```

```{r}
model2 <- lm(body_mass_g ~ bill_depth_mm) 
summary(model2)
```

```{r}
model3 <- lm(body_mass_g ~ flipper_length_mm) 
summary(model3)
```
For the first round of the forward model building, it seems all each predictor is significant (for estimation on population parameter $\beta$), since p-values are way smaller than 5% significance level. So, we choose with the most significant predictor among them and it is `fliper_length_mm` because its absolute t-value is highest (i.e., p-value is lowest).

Continue with model3, adding other predictors one by one.

2nd Round
```{r}
model4 <- lm(body_mass_g ~ flipper_length_mm + bill_length_mm)
summary(model4)
```

```{r}
model5 <- lm(body_mass_g ~ flipper_length_mm + bill_depth_mm)
summary(model5)
```

3rd Round
```{r}
model6 <- lm(body_mass_g ~ flipper_length_mm * bill_depth_mm)
summary(model6)
```

```{r}
model7 <- lm(body_mass_g ~ flipper_length_mm * bill_length_mm)
summary(model7)
```

```{r}
model8 <- lm(body_mass_g ~ flipper_length_mm + I(flipper_length_mm^2))
summary(model8)
```

4th Round
```{r}
model9 <- lm(body_mass_g ~ flipper_length_mm * bill_depth_mm + I(flipper_length_mm^2))
summary(model9)
```

```{r}
model10 <- lm(body_mass_g ~ flipper_length_mm * bill_depth_mm + flipper_length_mm * bill_length_mm)
summary(model10)
```

5th Round
```{r}
model11 <- lm(body_mass_g ~ flipper_length_mm * bill_depth_mm + flipper_length_mm * bill_length_mm + I(flipper_length_mm^2))
summary(model11)
```

```{r, echo=TRUE}
par(mfrow=c(2,2))
plot(model10)
```

```{r, echo=TRUE}
model12 <- lm(body_mass_g ~ flipper_length_mm * bill_depth_mm + flipper_length_mm * bill_length_mm + factor(species) + factor(sex) + factor(island))
summary(model12)
```

```{r, echo=TRUE}
model13 <- lm(body_mass_g ~ flipper_length_mm * bill_depth_mm + flipper_length_mm * bill_length_mm + factor(species) + factor(sex))
summary(model13)
```

```{r, echo=TRUE}
par(mfrow=c(2,2))
plot(model13)
```

```{r, echo=TRUE}
anova(model10, model13)
```