---
title: "Group12-Project3"
author: "Ganghyun Kim(01715699),Seungchan Oh(01603277), Wonjong Bum(01815150)"
date: '2022 5 21'
output: html_document
---

```{r pressure, echo=FALSE}
data <- read.csv("vanrolleghem_our.csv", header=TRUE)
attach(data)
source("monod-functions.R")
```

## 1. Read in the given dataset and make a plot of the oxygen uptake rate with respect to time.

```{r}
plot(our.data ~ our.times, main = "Data plot", xlab='t', ylab='Oxygen Uptake Rate')
```

## 2. Perform a parameter estimation for the single and double Monod models, using the following initial parameter estimates. Report the optimized parameter values.

#### For **single** Monod:

> $\alpha = 0.5,\ \beta = 2.7,\ S(0) = 2.0$

#### For **double** Monod:

> $\alpha_1 = -1.20,\ \beta_1 = 0.34,\ S_1(0) = 1.39,\\ \alpha_2 = -0.69,\ \beta_2 = -1.61,\ S_2(0) = 0.34$

```{r}
# alpha, beta, S parameter
singlePar <- c(0.5, 2.7, 2.0) # Single
doublePar1 <- c(-1.20, 0.34,  1.39) # Double 1
doublePar2 <- c(-0.69, -1.61, 0.34) # Double 2 

# Objective function
Jobjective <- function(thetas, FUN, xdata, ydata) {
  y_est <- do.call(FUN, list(xdata, thetas))
  y_resid <- ydata - y_est 
  J <- sum(y_resid^2) 
  return(J)
}

# Optimization of models
singlefit <- optim(par = singlePar, fn = Jobjective, FUN = single.monod, xdata = our.times, ydata = our.data)
doublefit <- optim(par = c(doublePar1, doublePar2), fn = Jobjective, FUN = double.monod, xdata = our.times, ydata = our.data)
```

The optimized set of parameters for the single Monod model is,
```{r}
singlefit$par
```

The optimized set of parameters for the double Monod model is,
```{r}
doublefit$par
```

## 3. Plot the fitted models on one and the same plot.
```{r}
color <-  c('#805FFA', '#5DFA61')
xseq <- seq(min(our.times), max(our.times), 0.1)

plot(our.data ~ our.times, main = "Fitted model", xlab='t', ylab='Oxygen Uptake Rate')
lines(x=xseq, y=single.monod(xseq, singlefit$par), col = color[1])
lines(x=xseq, y=double.monod(xseq, doublefit$par), col = color[2])
legend(x = "topright", legend = c("single", "double"), fill = color)
```

## 4. Write a function in R to compute the Bayesian information criterion (BIC) and compare the goodness-of-fit using this criterion for both models. According to BIC, which is the better model? Does this agree with your intuition?

```{r}
# Bayesian Information Criterion (BIC) function is coded below.
BIC <- function(thetas, FUN, xdata, ydata) { #x y
  RSS <- Jobjective (thetas, FUN, xdata, ydata)
  N <- length(xdata)
  p <- length(thetas)
  B <- N*log(RSS/N) + p*log(N) # Akaike's: 2p, Bayesian: plog(N)
  return(B)
}

# Calculated BIC values for each model.
BIC(singlefit$par, single.monod, our.times, our.data) #The better
BIC(doublefit$par, double.monod, our.times, our.data)
```

 In order to pick the best model, perform Bayesian Information Criterion(BIC) which is an estimator of prediction error.
Since BIC indicates goodness-of-fit of each model, it provides a benchmark for model selection. BIC of double Monod is lower than one of single Monod. Therefore, the double Monod model is better description.

 In order to statistically verify which model is better, F-test is performed.

```{r}
Ftest <- function(thetas1, thetas2, FUN1, FUN2, xdata, ydata) {
  RSS1 <- Jobjective(thetas1, FUN1, xdata, ydata)
  RSS2 <- Jobjective(thetas2, FUN2, xdata, ydata)
  N <- length(xdata)
  p1 <- length(thetas1)
  p2 <- length(thetas2)
  F <- ((RSS1 - RSS2) / (p2 - p1)) / (RSS2 / (N - p2))
  pvalue <- pf(F, df1 = p2 - p1, df2 = N - p2, lower.tail = FALSE)
  return(list(F = F, df1 = abs(p2 - p1), df2 = N - p2, pValue = pvalue))
}
# F-test between single Monod model and double Monod model.
Ftest(singlefit$par, doublefit$par, single.monod, double.monod, our.times, our.data)
```

> $H_{0}$: The complex double Monod is not significantly better than simple single Monod. \
> $H_{A}$: The complex double Monod is significantly better than simple single Monod.

+ Critical F-value: 402.7574

+ P-value: 6.433483e^(-37)

 $H_{0}$ is rejected so that $H_{A}$ is accepted: double Monod is significantly better than single Monod. Even though double Monod is more complex than single Monod, the model is more effective.

## 5. Perform a local sensitivity analysis for each of the parameters of your selected model and make a single plot with the sensitivity functions for each parameter.
```{r}
relativeSensitivity <- function(FUN, thetas, xdata = 0:45, eps = 1e-5) {
	yhat <- do.call(FUN, list(xdata, thetas))
	S <- matrix(NA, nrow = length(xdata), ncol = length(thetas))
	for (p in 1:length(thetas)) {
		thetas. <- thetas
		thetas.[p] <- thetas[p] * (1+eps)
		yhat. <- do.call(FUN, list(xdata, thetas.))
		S[, p] <- ((yhat. - yhat) / (eps*thetas[p])) * abs(thetas[p]) #Sensitivity with respect to parameters.
	}
	
	return(S)
}

xseq <- seq(min(our.times), max(our.times), length.out=200)

SensArray <- relativeSensitivity(double.monod, doublefit$par, xdata = xseq,
                                 eps = 1e-5)
```

 To compare sensitivities of each parameter, relative sensitivity is used. Relative sensitivity with respect to parameter is  $\frac{\partial y_i(t)}{\partial \theta_j} \cdot \theta_j$.

```{r}
plot(xseq, SensArray[,1],  main = "Sensitivity for each parameter in the double Monod", type = "l", lty=3, frame = TRUE, pch = 18,
     col = "red", ylim = c(-3,4), xlab = "t",
     ylab = "Sensitivity")
lines(xseq, SensArray[,2], pch = 18, col = "#F07F00", type = "l", lty = 3)
lines(xseq, SensArray[,3], pch = 18, col = "#D9F50C", type = "l", lty = 3)
lines(xseq, SensArray[,4], pch = 18, col = "#EB0C82", type = "l", lty = 3)
lines(xseq, SensArray[,5], pch = 18, col = "#0C1AF5", type = "l", lty = 3)
lines(xseq, SensArray[,6], pch = 18, col = "#16DE98", type = "l", lty = 3)
legend(x = "topright", legend = c(expression(alpha[1]), expression(beta[1]), expression(S[1]), expression(alpha[2]), expression(beta[2]), expression(S[2])), fill = c("red", "#F07F00", "#D9F50C", "#EB0C82", "#0C1AF5", "#16DE98"))
```
```{r}
par(mfrow=c(2, 3))
plot(xseq, SensArray[,1],  main = expression(paste("Sensitivity of ", alpha[1])), type = "l", frame = TRUE, pch = 18,
     col = "red", ylim = c(-3,4), xlab = "t",
     ylab = "Sensitivity")
plot(xseq, SensArray[,2],  main = expression(paste("Sensitivity of ", beta[1])), type = "l", frame = TRUE, pch = 18,
     col = "#F07F00", ylim = c(-3,4), xlab = "t",
     ylab = "Sensitivity")
plot(xseq, SensArray[,3],  main = expression(paste("Sensitivity of ", S[1])), type = "l", frame = TRUE, pch = 18,
     col = "#D9F50C", ylim = c(-3,4), xlab = "t",
     ylab = "Sensitivity")
plot(xseq, SensArray[,4],  main = expression(paste("Sensitivity of ", alpha[2])), type = "l", frame = TRUE, pch = 18,
     col = "#EB0C82", ylim = c(-3,4), xlab = "t",
     ylab = "Sensitivity")
plot(xseq, SensArray[,5],  main = expression(paste("Sensitivity of ", beta[2])), type = "l", frame = TRUE, pch = 18,
     col = "#0C1AF5", ylim = c(-3,4), xlab = "t",
     ylab = "Sensitivity")
plot(xseq, SensArray[,6],  main = expression(paste("Sensitivity of ", S[2])), type = "l", frame = TRUE, pch = 18,
     col = "#16DE98", ylim = c(-3,4), xlab = "t",
     ylab = "Sensitivity")
```
 
 Sensitivity of all parameters eventually converges to 0 as time goes. Two parameters, $S_2$ and $\alpha_2$, are the most sensitive among parameters in the model and each pulses positively or negatively at rough time 10, respectively. The sensitivity of $\beta_2$ also pulses at the same time, but not much as former two.
