---
output: html_document
---

# Exercise 15

For each test you perform in this exercise...

...you give a figure or table which should give you a comparison "at first glance"
...you justify the choice for the test
...you give the null and alternative hypothesis
...you give the test assumptions and you check them where possible
...you write a conclusion based on the output

The dataset immer.dat contains data on the yield of five barley varieties grown at six locations in 1931 and 1932 (Immer et al., 1934). The dataset contains the following variables:

* Loc: The location.
* Var: The variety of barley (M=”manchuria”, S=”svansota”, V=”velvet”, T=”trebi” and P=”peatland”).
* Y1: Yield in 1931. 
* Y2: Yield in 1932.

read the data
```{r}
imm<-read.table("immer.dat", header=T, sep="\t")
attach(imm)
```

### Question 1:
### We wish to examine whether there is a difference in the yield between 1932 and 1931 among all the varieties and locations. Make a graph which gives you insight into the research question.

graphical comparison: boxplot:
```{r}
boxplot(cbind(Y1,Y2) ~ Loc + Var)
```


### Question 2:
### Which statistical test would you choose here? Argue why.

Anova test
Reason : comparison of yields between different groups

### Question 3:
### Formulate the null and alternative hypotheses for this test.

* $H_0$: The yield is the same in all groups
* $H_A$: The yield differs between at least 2 groups.


### Question 4:
### Perform the test at the 5% significance level, report the p-value and formulate a conclusion.

```{r}
imm.aov_LocVar <- aov(cbind(Y1,Y2) ~ Loc + Var, data = imm)
summary(imm.aov_LocVar)
```

```{r}
imm.aov_LocVarM <- aov((Y1+Y2)/2 ~ Var + Loc, data = imm)
summary(imm.aov_LocVarM)
```
### Question 5:
### Suppose one wishes to know if the yield of 1931 is different between locations, ignoring the varieties. Make a graph. Which statistical test would you choose now? Perform the test at the 5% significance level, report the p-value and formulate a conclusion.

```{r}
boxplot(cbind(Y1,Y2) ~ Loc)
```

Anova test
Reason : comparison of yields between different groups

```{r}
imm.aov_Loc <- aov(cbind(Y1,Y2) ~ Loc, data = imm)
summary(imm.aov_Loc)
```

```{r}
imm.aov_LocM <- aov((Y1+Y2)/2 ~ Loc, data = imm)
summary(imm.aov_LocM)
```


```{r}
detach(imm)

```