---
output: html_document
---

# Exercise 14

For each test you perform in this exercise...

...you give a figure or table which should give you a comparison "at first glance"
...you justify the choice for the test
...you give the null and alternative hypothesis
...you give the test assumptions and you check them where possible
...you write a conclusion based on the output

The dataset cardiac.dat was obtained in a study where it was determined if the administration of the substance dobutamine can be used to predict cardial attacks. This substance was administered to 588 random persons and causes an extra load on the heart. The following variables where then measured to be able to predict if the person could have a risk for a cardial attack or not:

* dose: dose of dobutamine 
* maxhr: maximal heart rate after administration of dobutamine 
* age: age of the patient 
* gender: gender of the patient (0 = M, 1 = F)
* smoking: 0 = smoker, 0.5 = occasional smoker, 1 = not a smoker
* any.event: 0 = patient would get a heart attack within the year after the medical examination, 1 = patient would not get a heart attack within the year after the medical examination

read the data
```{r}
cardiac<-read.table("cardiac.dat", header=T, sep=" ")
attach(cardiac)
```


### Question 0 :
### First test if there is a difference in the variance of the maximal heart rate for men and women. Report the p-value and give a conclusion. Give a figure which supports the conclusion.

equal variances?

* $H_0$: ...
* $H_A$: ...

```{r}

```

...

### Now test if the mean maximal heart rate differs between men and women. Take into account the previous result for your choice. 

Comparing the mean between two independent groups => unpaired t-test
with equal variances

* $H_0$: ...
* $H_0$: ...

graphical comparison: boxplot
```{r}

```

based on the boxplot you can expect ...


Conditions:

- ...
- ...
- ...

```{r}
# QQNorm & QQLine



```

...

Test:
```{r}
# Run t-test

```

There is no proof that the heart rate significantly differs (as expected from the boxplot)
p=0.8335

### Question 1-3
### Question 1: One wishes to examine if there is a difference in the mean maximal heart rate depending on the smoking status. If you could choose 1 statistical test to examine this, which would you choose? Argue why.
### Question 2: Perform the test at the 5% significance level, report the p-value and formulate a concrete and insightful conclusion.
### Question 3: Which assumptions need to be satisfied to perform this test? Verify graphically if these are fullfilled and describe which plots you made and what you see on them.

comparison of the means between 3 groups => Anova

* $H_0$: ...
* $H_A$: ...

graphical comparison: boxplot
```{r}
boxplot(maxhr~smoking)
```

conditions:

- ...
- ...
- ...
- ...

```{r}
# QQNorm & QQLine

```

...

```{r}
table(smoking)
boxplot(maxhr~smoking)
```

...


(variances could be tested at the 5%/3=1.67% significance level)

```{r}
# Run Var test for each of smoking category

```

...

Actual test:
```{r}

```

...

(Or using a non-parametric alternative because of non-normality)
```{r}
# Run Krushal Test


```

### Question 4
### Suppose then you wishes to examine if there is an association between smoking status and gender. Which statistical test can you use? Argue why.

two categorical variables => chi-square test

* $H_0$: ...
* $H_A$: ...

contingency table
```{r}

```

At first glance: ...

### Question 5:
### Perform the test at the 1% significance level, report the p-value and formulate a conclusion.

```{r}
# Run Chi-squared test


```

...

### Question 6:
### Investigate if the assumptions are fullfilled to perform this test.

Conditions:

- ...
- ...


```{r}
# Show expected data


```

...


### Question 7-9
### Question 7: Report (a) the relative risk of being a smoker (regularly or occasionally) for men versus women, and (b) the relative risk of being an occasional smoker for smoking men versus smoking women. 
### Question 8: Calculate for both relative risks the corresponding confidence interval, and make sure the overall confidence level is (approximately) 95% (so make sure the probability for at least 1 of the intervals to not contain the relative risk of the population, is 5% at the most).
### Question 9: Interpret the results from the above two research questions.

a)
```{r}






```

Interpretation: ...

b)
```{r}




 

```

...

