---
title: "PC Practical: Linear regression II"
output: html_document
---

## Exercise 17

The data in the dataset calibration.txt were collected by students at the VIB in a study on the saccharification potential of a wild-type poplar and a transgene CCR poplar. Cellulose is first decomposed by cellulase and beta-glucosidase. The released glucosis in the hydrolised samples is then estimated using a combined enzymatic test, i.e., glucose oxidase type II-S from _Aspergillus niger_ (GOD) and
peroxidase type II from horseradish (POD) with the chromogen ABTS (2,2’-azino-bis(3-ethylbenzothiazoline-6-sulphonic acid)). The accompanying reactions are for 

__GOD: D-glucose + O~2~ <---> D-gluconic acid + H~2~O~2~ __

and for 

__POD: H~2~O~2~ + ABTS <---> ABTS^+^ + H~2~O. __

GOD catalyses the oxidation of glucose during which H~2~O~2~ is released. This H~2~O~2~ oxides ABTS in the presence of POD to the green radical cation ABTS^+^. The increase in absorbance for 405 nm was measured for known concentrations of glucose.

Read the data via `read.table`. Change your working directory to the folder where you stored `calibration.txt`. 

```{r} 
calib <- read.table("calibration.txt", sep="", header = TRUE) 
attach(calib)
```

#### 1. Fit a linear regression model for the absorbance for 405 nm (`A405`) in function of the percentage glucose (`gluc_percent`). Interpret the result.

```{r}
model <- lm(A405 ~ gluc_percent)

plot(A405 ~ gluc_percent)

summary(model)
confint(model)
```

There is a highly significant effect ($p< 0.001$) of the percentage glucose on
the absorbance for 405 nm. The absorbance for 450 nm is estimated to increase
with 3.017 (95\% CI [2.925, 3.108]), for every increase in the percentage
glucose of 0.001.

#### 2. Compose a calibration curve together with the accompanying 95\% confidence intervals. Do the same, but with 95\% prediction intervals. Interpret the results.

Since we want to compose a calibration curve, we want to have predictions for
each $X$-value. Thus first create a sequence of new data points to predict and
then predict the outcome using the linear model. 

Take care, when predicting, make sure that either your new data are in a
dataframe with column name 'gluc_percent' or specifiy this in your prediction
step as below.

```{r}
newx <- seq(min(gluc_percent), max(gluc_percent), length = 1000)
CI <- predict(model, newdata = data.frame(gluc_percent = newx), interval = 'confidence')
PI <- predict(model, newdata = data.frame(gluc_percent = newx), interval = 'prediction')
```

Next plot the curve with confidence and prediction intervals:

```{r, fig.cap = "Figure 1: ..."}
plot(A405 ~ gluc_percent, type = "n")
polygon(c(rev(newx), newx), c(rev(PI[, 3]), PI[, 2]), col = "grey90", border = NA)
lines(newx, PI[, 3], lty = "dashed")
lines(newx, PI[, 2], lty = "dashed")
polygon(c(rev(newx), newx), c(rev(CI[, 3]), CI[, 2]), col = "grey75", border = NA)
lines(newx, CI[, 3], lwd = 1)
lines(newx, CI[, 2], lwd = 1)
abline(model, lwd = 2)
```

The 95\% confidence interval contains with 95\% probability the true regression
line. 

The 95\% prediction interval contains with 95\% probability a random observation
of the absorbance for 405 nm, given a certain value for the predictor variable
percentage glucose.

#### 3. Verify whether the model describes the data properly. If not, adapt
the model and the obtained calibration curves.

```{r}
par(mfrow=c(2,2))
plot(model)
```

The linearity assumption does not seem to hold. Also normality and
homoscedasticity do not seem to be OK. To solve linearity we can add a quadratic
term.

```{r}
model2 <- lm(A405 ~ gluc_percent + I(gluc_percent^2))
summary(model2)
anova(model, model2)

par(mfrow = c(2,  2))
plot(model2)
```

Indeed adding the quadratic term significantly contributes in explaining the
variance in the absorbance for 450nm given that the percentage glucose is
already in the model.

Linearity: maybe, but let's try to improve it by adding a cubic term

```{r}
model3 <- lm(A405 ~ gluc_percent + I(gluc_percent^2) + I(gluc_percent^3))
summary(model3)
anova(model2, model3)

par(mfrow = c(2, 2))
plot(model3)
```

Slight improvement for linearity, but normality became worse. Adding the cubic
term significantly contributes in explaining the variance in the absorbance for
450nm given that there is already a first and second order term for percentage
glucose in the model, but it is only just significant

What if we add a fourth order term?

```{r}
model4 <- lm(A405 ~ gluc_percent + I(gluc_percent^2) + I(gluc_percent^3) + I(gluc_percent^4))
summary(model4)
anova(model3, model4)

par(mfrow = c(2, 2))
plot(model4)
```

It seems like the fourth order term is not needed. So we continue with the model
of degree 3. We'll try some transformations to improve the normality and
homoscedasticity.

```{r}
model3log <- lm(log(A405) ~ gluc_percent + I(gluc_percent^2) + I(gluc_percent^3))
summary(model3log)

par(mfrow = c(2, 2))
plot(model3log)
```

```{r}
model3inv <- lm(1/A405 ~ gluc_percent + I(gluc_percent^2) + I(gluc_percent^3))
summary(model3inv)

par(mfrow = c(2, 2))
plot(model3inv)
```

```{r}
model3sqrt <- lm(sqrt(A405) ~ gluc_percent + I(gluc_percent^2) + I(gluc_percent^3))
summary(model3sqrt)

par(mfrow = c(2, 2))
plot(model3sqrt)
```

We choose the 3rd order model with sqrt-transformation.

```{r}
newx <- seq(min(gluc_percent), max(gluc_percent), length = 1000)
CI <- predict(model3sqrt, newdata = data.frame(gluc_percent = newx), interval = 'confidence')^2
PI <- predict(model3sqrt, newdata = data.frame(gluc_percent = newx), interval = 'prediction')^2

plot(A405 ~ gluc_percent, type = "n")
polygon(c(rev(newx), newx), c(rev(PI[, 3]), PI[, 2]), col = "grey90", border = NA)
lines(newx, PI[, 3], lty = "dashed")
lines(newx, PI[, 2], lty = "dashed")
polygon(c(rev(newx), newx), c(rev(CI[, 3]), CI[, 2]), col = "grey75", border = NA)
lines(newx, CI[, 3], lwd = 1)
lines(newx, CI[, 2], lwd = 1)
points(A405 ~ gluc_percent)
```

## Exercise 18
In a study of Lewis (Lewis et al., 1937) that basal metabolism of 50 different boys
was measured in calories per hour while they were lying still on a bed in a closed
room. Their ages varied from 32 to 152 months. Also their weight and height were
measured.

The dataset metadata.txt contains thus the variables:

* __age__: age in month
* __weight__: weight in kg
* __height__: height in cm
* __meta__: metabolism in calories per hour (on logarithmic scale (-1))


```{r} 
metadata <- read.table("metadata.txt", sep="", header = TRUE) 
attach(metadata)
```

To not make interpretations needlessly complex: `meta' is the metabolism in
calories per hour on a logarithmic scale, and without the factor "-1".

#### 1. Find a best model to predict the mean metabolism in function of weight, height and age.

**Use forward selection to find the optimal model**

```{r} 

```

...

What about quadratic effects?

```{r} 
model7 <- lm(meta ~ height + weight + age + I(age^2)); summary(model7)
```

...

```{r}
model8 <- lm(meta ~ weight + age + I(age^2)); summary(model8)
```

Are there other significant quadratic effects?

```{r}
model9 <- lm(meta ~ weight + I(weight^2) + age + I(age^2)); summary(model9)
```

...

Are there significant interactions?

```{r}
modelB <- lm(meta ~ weight + age + I(age^2) + height:age + height:I(age^2))
summary(modelB)
modelC <- lm(meta ~ weight + age + I(age^2) + height:age)
summary(modelC)
```

...

#### 2. Interpret the estimated parameters in your final model, and interpet the multiple correlation coefficient.

```{r} 
finalmodel <- model8
summary(finalmodel)
```

- a. Interpretation of the weight parameter estimate:

...

- b. Interpretation of the age parameter estimate:

The quadratic term makes it difficult to interpret the main effect and the
quadratic effect of age separately, but we can interpret the terms together.

   - Consider two ages $l$ and $l+1$.
   - $\hat{\beta_1}$ is the estimated main effect of age.
   - $\hat{\beta_2}$ is the estimated quadratic effect of age.
   - $\hat{\beta_2}$ is the estimated effect of weight.
   
When we keep the predictor variable "weight" constant at the value $g$, then:

$\widehat{meta(l)} = \hat{\beta_2} l^2 + \hat{\beta_1} l + \hat{\beta_3} g$

and

$\widehat{meta(l+1)} = \hat{\beta_2} (l+1)^2 + \hat{\beta_1} (l+1) + \hat{\beta_3} g$

$= \hat{\beta_2} (l^2+2l+1) + \hat{\beta_1} (l+1) + \hat{\beta_3} g$

$= \hat{\beta_2} l^2 + 2 \hat{\beta_2} l+ \hat{\beta_2} + \hat{\beta_1} l + \hat{\beta_1} + \hat{\beta_3} g$

When we substract these from each other, we get:

$\widehat{meta(l+1)} - \widehat{meta(l)} = (\hat{\beta_2} l^2 + 2 \hat{\beta_2} l+ \hat{\beta_2} + \hat{\beta_1} l + \hat{\beta_1} + \hat{\beta_3} g) - (\hat{\beta_2} l^2 + \hat{\beta_1} l + \hat{\beta_3} g)$

$= \hat{\beta_2} l^2 + 2 \hat{\beta_2} l+ \hat{\beta_2} + \hat{\beta_1} l + \hat{\beta_1} + \hat{\beta_3} g - \hat{\beta_2} l^2 - \hat{\beta_1} l - \hat{\beta_3} g$

$= 2 \hat{\beta_2} l+ \hat{\beta_2} + \hat{\beta_1}$

So the increase depends on the value of the "age" variable $l$.

If we plug in the values $\hat{\beta_1} = 2.786*10^{-03}$ and
$\hat{\beta_2} = -1.312*10^{-05}$, we get:

$\widehat{meta(l+1)} - \widehat{meta(l)} = -2.624*10^{-05} l - 1.312*10^{-05} + 2.786*10^{-03}$

$= -2.6*10^{-05} l + 0.0028$

Interpretation:

When you compare persons who differ in age by one year, but which have the same
weight, the mean logarithm of the metabolism is estimated to increase by 0.0028
minus 2.6e-5 times the age of the persons in the youngest group.

**Interpretation on the original scale:**

...

**Interpret the multiple correlation coefficient** 

...


#### 3. Make appropriate QQ-plots and residual plots to assess the final model
fit. Interpret the figures and explain for each figure what any possible
deviation represents. 

```{r}
par(mfrow=c(2,2))
plot(finalmodel)
```

Interpretation:

1. Upper left: ...
   
2. Lower left: ...
   
3. Upper right: ...
   
4. Lower right: This graph shows the standardised residuals plotted against the
   leverage. There are three important things you can deduce from this graph:
    1. The leverage is a measure for the potential of the data point to
       influence the parameter estimates, if the observation would be shifted in
       the y-direction by one unit. It indicates which observations are extreme
       in terms of the predictors. The leverage is considered extreme if it is
       higher than $2p/n = 2*4/50 = 0.16$. There are quite some points which
       have an extreme leverage.
    2. (Standardized) residuals indicate how extreme observations are in terms
of the response variable. Residuals are considered extreme if the studentized
residuals (a similar transformation as standardized residuals) surpasses
$1.96*\hat{\sigma}$, where $\hat{\sigma}$ is the residual standard deviation.
    3. This graph also shows the contours for the Cook's distance equal to 0.5
and 1. Points with a high Cook's Distance have a high influence on the estimated
coefficients and so also on the predictions. They are considered extreme if they
       pass the 50\% percentile of a $F_{p,n-p}$-distribution. The 50\%
       percentile of a $F_{4,46}$-distribution is equal to 0.85. There are no
       points with an extreme Cook's Distance. 

#### 4. Investigate if the data of certain boys have a large influence on all the fitted values and coefficients. If necessary adjust the model. 

Do the data of certain boys have and extreme influence on all the fitted values
and coefficients: Cook's Distance (interpretation, see last question):

```{r}
par(mfrow=c(1,1))
cook_final <- cooks.distance(finalmodel)
plot(cook_final,type='h')
p <- length(coef(finalmodel))
n <- nrow(metadata)
fquantile <- qf(0.5,df1=p,df2=n-p) #Threshold for Cook's Distance
all(cook_final < fquantile) #TRUE => All Cook's Distances are smaller then the 50% percentile of $F(p,n-p)$-distribution
plot(cook_final,type='h',ylim=c(0,1))
abline(h=qf(0.5,df1=p,df2=n-p))
```

Do the data of certain boys have an extreme infuence on certain coefficients?
```{r}
#dfbetas
dfbeta_finalmodel <- dfbetas(finalmodel)
treshold <- 2 / sqrt(n) #Relatively large dataset, we can use 2 / sqrt(n) as the critical value for DFBETAS
all(dfbeta_finalmodel > treshold) #FALSE => some DFBETAS are larger than the critical value
which(dfbeta_finalmodel[,"(Intercept)"] > treshold) #intercept
which(dfbeta_finalmodel[,"weight"] > treshold) #weight
which(dfbeta_finalmodel[,"age"] > treshold) #age
which(dfbeta_finalmodel[,"I(age^2)"] > treshold) #age^2
```

What we did here was to check, for each parameter separately, which observations
have an extreme influence on that parameter. We verify if some observations
reoccur several times, because this would mean they have a strong influence on
multiple parameters. Observations 25 and 50 have a strong influence on both the
intercept and the quadratic effect of age.

```{r}
metadata[c(25,50),]
```

What if we fit the same model without these observations?

```{r}
metadata2 <- metadata[-c(25,50),]
modelWithout <- lm(metadata2$meta ~ metadata2$weight + metadata2$age + I(metadata2$age^2))
summary(modelWithout)
summary(model8)
```

The differences are small. We leave these observations in. What we did here is
dangerous. In general it is a bad idea to just throw away data, unless you are
very sure that the data are wrong, or not relevant for some  reason. So don't
just throw away data!

Verify if those two observations are outliers when weight is plotted in function
of age, age in function of the logarithm of the metabolism and weight in
function of the logarithm of the metabolism:

```{r}
colors <- rep(1,50)
colors[c(25,50)] <- 2
plot(x=weight,y=age,col=colors)
plot(x=age,y=meta,col=colors)
plot(x=weight,y=meta,col=colors)
```

#### 5. Is there multicollinearity in the final model? If necessary adjust the
model.

Remark: If you would have added a quadratic effect of height (which is slightly
more significant) instead of a quadratic effect of age, then you would have
ended with the following model:

```{r}
modelD <- lm(meta ~ height + weight + I(height^2))
summary(modelD)
```

If you wouldn't have added age as a quadratic effect, you could have ended with
the following model:

```{r}
modelE <- lm(meta ~ height + weight + height:weight)
summary(modelE)
```

If you would do backward selection (assuming we don't add a quadratic trend for
the logarithm of the metabolism in terms of weight and lenght):

```{r}
modelFull <- lm(meta ~ height + weight + age + I(age^2) + height:weight + height:age + height:I(age^2) + weight:age + weight:I(age^2))
```

```{r}
step <- lm(meta ~ height + weight + age + height:age)
summary(step)
```

The order in which we remove terms (do this yourself as an exercise!):
height:age^2, height:weight, weight:I(age^2), I(age^2), weight:age. Remark that
you should never remove main effects when their corresponding quadratic or
interaction terms are still in the model! The intercept should always stay in
the model!

There are a lot of different solutions. Why is this? What is wrong?

```{r}
cor(height, weight)
cor(height,age)
cor(weight,age)
```

Aha...

```{r}
library(faraway)
vif(model8)
vif(modelE)
vif(step)
```

VIF stands for the Variance Inflation Factor. The VIF for the k-th coefficient
in the regression model is defined as: $VIF = (1-R_k^2)^{-1}$

Here $R_k^2$ is the multiple correlation coefficient of a linear regression with
the k-th predictor on all other predictors in the model. The VIF has as a
property that she is equal to 1 if the k-th predictor is not linearly associated
with th eother predictors in the model (so when the k-the predictor is perfectly
uncorrelated with the other predictors). The VIF is larger than 1 in all other
cases. The VIF also expresses how much larger the observed variance on the
estimation of the k-the coefficient is compared with when all predictors would
be independent. The larger the VIF, the less stable the estimates will be. We
talk about strong multicollinearity when the VIF is larger than 10.

We notice an extremely high multicollinearity!

As the height is explaining most of the variance in the response variable, we
will propose a model with only this variable. To do: validate this model again
(residual plots, Cook's Distance, ...).

```{r}
#Old model:
summary(finalmodel)
#New model:
finalmodel2 <- lm(meta ~ height)
summary(finalmodel2)
```

**Compare the $R^2$ of the last "final" model. What do you see?**

...

Another option would be to center the predictors around zero (this is advised
when you include interaction and quadratic term in the model).

```{r}
c.height <- scale(height,center=TRUE,scale=FALSE)
c.weight <- scale(weight,center=TRUE,scale=FALSE)
c.age <- scale(age,center=TRUE,scale=FALSE)
model7B <- lm(meta ~ c.height + c.weight + c.age + I(c.age^2))
```

```{r}
summary(model7B)
vif(model7B)
```

The VIF is now smaller for some of the effects, but still too high. Centering
doesn't improve much in this case, but it can be a solution in other cases. It
is always worth trying!
